### 2023-02-12 doc/manuscript/translocation.qmd
Draft paragraph 1 and 2 of introduction

P1: Emerging infectious diseases pose a severe threat to human, livestock, and wildlife populations.

-   In wildlife populations, disease has been responsible for dramatic declines, extirpations, and even extinctions in mammals, birds, echinoderms, amphibians, and other taxa (Scheele, avian malaria, black-footed ferrets, sea stars, etc.).

-   Although the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting, understanding, and mitigating disease impacts (citations), populations previously decimated by disease are also beginning to recover (citations).

-   However, given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery and how these mechanisms can be leveraged to re-establish host populations across large spatial scales.

-   In particular, for species of conservation concern as a result of disease-induced declines, demonstrating the long-term viability and feasibility of potential recovery actions in the presence of disease and elucidating the mechanisms underlying long-term recovery success remain largely absent from conservation biology (but see XX).

P2: Host evolution following disease-induced declines is one mechanism by which hosts can recover (Carlson et al. 2014, TREE, Searle et al. 2020).

-   Evolutionary rescue describes natural host evolution whereby rapid evolutionary changes originating from variation within a host population reverse ongoing population declines (Carlson).

-   In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes (Golas et al. 2021).

-   Evolutionary rescue can be difficult to detect, and this complicates our understanding of the potential importance of evolutionary rescue in allowing the persistence and recovery of host populations following disease emergence (Eskew et al. 2021, Lilley et al. 2020). <!--# Maybe move this to the Discussion? -->

-   Important for conservation action, evolutionary rescue not only promotes recoveries of focal populations (Gignous-Wolfsohn et al. 2021), but can also provide sources of resistant individuals to reestablish populations extirpated following disease emergence (Mendelson et al. 2019, Joseph and Knapp 2018).

@@

>>>>

Emerging infectious diseases pose a severe threat to human, livestock, and wildlife populations. In wildlife populations, disease has been responsible for dramatic declines, extirpations, and even extinctions in mammals, birds, echinoderms, amphibians, and other taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021]. Although the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting, understanding, and mitigating disease impacts [@langwig2015], populations previously decimated by disease are also beginning to recover [@newell2013; @voyles2018]. Given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery [@brannelly; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019]. In particular, for species of conservation concern as a result of disease-induced declines, demonstrating the long-term viability and feasibility of potential recovery actions in the presence of disease and elucidating the mechanisms underlying long-term recovery success remain largely absent from conservation biology [@venesky2012; @mendelson2019].

Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020]. Evolutionary rescue describes natural host evolution whereby rapid evolutionary changes originating from variation within a host population reverse ongoing population declines [@carlson2014]. In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021]. Evolutionary changes following disease-induced declines have been observed in black-lipped abalone populations suffering from herpes virus [@holland2022], little brown bat populations suffering from white-nose syndrome [@gignoux-wolfsohn2021], Tasmanian devils suffering from facial tumor disease [@epstein2016], and amphibian populations suffering from chytridiomycosis [@savage2016]. However, linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, remains rare. Important for conservation action, selection that promotes recoveries of focal populations can also provide sources of resistant individuals to reestablish populations extirpated following disease emergence [@joseph2018; @mendelson2019]. These so-called "recovery engines" [@mendelson2019] -- where rescued (i.e., resistant and recovered) host populations are used to successfully repatriate other extirpated populations on the landscape -- are an intriguing management idea, but have not been successfully implemented in wildlife conservation efforts.

@@

<!--#  Suggested start to this paragraph to link with the previous paragraph: "There are four conditions that will often..." -->
